

# Transition matrix

```{r define transition matrix, message = FALSE}
library(mstate)

# Define transition matrix ------------------------------------------------

# Simple illness death transition matrix

tmat <- transMat(x = list(c(2, 3,  6, 7), 
                          c(4, 6, 7), 
                          c(5, 6, 7), 
                          c(6, 7),
                          c(6, 7), 
                          c(),
                          c()), 
                 names = c("Healthy", 
                           "INT", 
                           "CMD", 
                           "INT->CMD", 
                           "CMD->INT",
                           "CVE",
                           "Death"))


```

# Prepare the data for multi state


```{r prepare data for analysis, warning=FALSE}
# Prepare the data for analysis -------------------------------------------

# Define covariates to keep

covs <- c("gender","ethnicity", "smoking_status","baseline.a","int.a","cmd.a","int.cmd.a","cmd.int.a","cve.a","death.a","IMD","icmm_pgs")



# prepare the data
reference_ms_df <- msprep(
  time = c(
    NA,
    "int.t",
    "cmd.t",
    "int.cmd.t",
    "cmd.int.t",
    "cve.t",
    "death.t"
  ),
  status = c(
    NA,
    "int.i",
    "cmd.i",
    "int.cmd.i",
    "cmd.int.i",
    "cve.i",
    "death.i"
  ),
  data = my_df,
  trans = tmat,
  keep = covs
)

```

# Number of events

```{r number of events}

# Check number of events

# Frequencies
kableExtra::kable(events(reference_ms_df)$Frequencies) %>%
  kableExtra::kable_styling()

# Proportions
kableExtra::kable(round(events(reference_ms_df)$Proportions*100, digits=2))%>%
  kableExtra::kable_styling()

```



# Knot fit check

```{r Knot fitting,eval = TRUE}

max_paths<-max(tmat, na.rm=TRUE)
max_knots <- 5
n_knots <- rep(NA,max_knots)
LL <- rep(NA,max_knots)
AIC <- rep(NA,max_knots)
BIC <- rep(NA,max_knots)      

knot_df <- lapply(1:max_paths,
                  function(x) {
                    for (i in 1:max_knots) {
                      the_model <- flexsurvspline(
                        Surv(time = time, event = status)~1,
                        # inits = c(gamma0 = 1, gamma1 = 0.1,gamma2=0.1),
                        subset = (trans == x),
                        data = reference_ms_df,
                        k = i - 1,
                        scale = "hazard"
                      )
                      
                      n_knots[i] <- i - 1
                      LL[i] <- the_model$loglik
                      AIC[i] <- the_model$AIC
                      BIC[i] <- the_model$BIC
                      
                   }
                   
                   
                   # Make the mod fit data frame
                   mod_fit <- data.frame(n_knots, 
                                         LL, 
                                         AIC, 
                                         BIC
                                         )
                   
            
                 })
           

names(knot_df)<-paste0("transition_",1:max_paths)


```

```{r plot out the knot fits, eval = TRUE}


lapply(seq_along(knot_df), function(x) {
  knot_df[[x]] %>% pivot_longer(-n_knots, names_to = "stat", values_to = "value") %>%
    ggplot(aes(x = n_knots, y = value)) +
    geom_point() +
    geom_line() +
    ggtitle(names(knot_df)[x])+
    facet_wrap( ~ stat, scales = "free_y")
  ggsave(paste0(names(knot_df)[[x]],"_plos_r1.png"), width = 10,height=4)
})


```
Optimum number of knots decided by visual inspection of AIC plots - aiming for lowest AIC OR clear inflection points in the graph

# Knot selection

```{r select number of knots}

t1_knots<-3
t2_knots<-2
t3_knots<-4
t4_knots<-1
t5_knots<-1
t6_knots<-2
t7_knots<-0
t8_knots<-2
t9_knots<-2
t10_knots<-2
t11_knots<-1
t12_knots<-0
t13_knots<-2
t14_knots<-1


```

# Transition 1: Healthy -> INT

```{r transition 1 rcs, eval = TRUE}
# Best fitting model number of knots
# t1_knots<-3


# Run the best fitting model adjusting for covariates using NPH from the cox models

t1_rcs <- flexsurvspline(
  Surv(time = time, event = status) ~
        rcs(baseline.a) + gender + ethnicity  + IMD + smoking_status+icmm_pgs ,
  anc = list(gamma0 ~ baseline.a+ smoking_status ,
             gamma1 ~ baseline.a+ smoking_status ,
             gamma2 ~ baseline.a+ smoking_status ,
             gamma3 ~ baseline.a+ smoking_status ,
             gamma4 ~ baseline.a+ smoking_status ),
  subset = (trans == 1),
  data = ms_df,
  k = t1_knots,
  scale = "hazard"
)
# Show the model
t1_rcs
t1_rcs%>%tidy(conf.int=TRUE, exponentiate=TRUE)
# Save the model
saveRDS(t1_rcs, file = "t1_rcs.rda")

```


```{r plot t1 models}

plot(t1_rcs,type = "cumhaz", ci = FALSE, main="Cumulative hazards")
plot(t1_rcs,type = "hazard", ci = FALSE, main = "Hazards", max.time=10)
plot(t1_rcs,type = "survival", ci = FALSE, main = "Survival")

```


# TRANSITION 2: Healthy -> CMD


```{r transition 2 rcs, eval = TRUE}
# Best fitting model number of knots


t2_rcs <- flexsurvspline(
  Surv(time = time, event = status) ~ 
    rcs(baseline.a) + gender + ethnicity  + IMD + smoking_status +icmm_pgs,
  anc = list(gamma0 ~ baseline.a+gender+ethnicity,
             gamma1 ~ baseline.a+gender+ethnicity,
             gamma2 ~ baseline.a+gender+ethnicity,
             gamma3 ~ baseline.a+gender+ethnicity),
  subset = (trans == 2),
  data = ms_df,
  k = t2_knots,
  scale = "hazard"
)


# Save the model
saveRDS(t2_rcs, file = "t2_rcs.rda")
```

```{r t2 load model}

plot(t2_rcs,type = "cumhaz", ci = FALSE, main="Cumulative hazards")
plot(t2_rcs,type = "hazard", ci = FALSE, main = "Hazards", max.time=10)
plot(t2_rcs,type = "survival", ci = FALSE, main = "Survival")

```

#  Transition 3: Healthy -> CVE

```{r transition 3 rcs, eval = TRUE}
# Best fitting model number of knots

t3_rcs <- flexsurvspline(
  Surv(time = time, event = status) ~
   rcs(baseline.a) + gender + ethnicity +  IMD+ smoking_status+icmm_pgs ,
  subset = (trans == 3),
  data = ms_df,
  k = t3_knots,
  scale = "hazard"
)

t3_rcs
# Save the model
saveRDS(t3_rcs, file = "t3_rcs.rda")

```

```{r t3 load model}
# load the model

plot(t3_rcs,type = "cumhaz", ci = FALSE, main="Cumulative hazards")
plot(t3_rcs,type = "hazard", ci = FALSE, main = "Hazards", max.time=10)
plot(t3_rcs,type = "survival", ci = FALSE, main = "Survival")

```

# Transition 4: Healthy -> Death

```{r transition 4 rcs, eval  =TRUE}

t4_knots
# Run the best fitting model
t4_rcs <- flexsurvspline(
  Surv(time = time, event = status) ~ 
    baseline.a + gender + ethnicity + IMD+  smoking_status +icmm_pgs,
  subset = (trans == 4),
  data = ms_df,
  k = 0,
  scale = "hazard"
)


saveRDS(t4_rcs, file = "t4_rcs.rda")

```

```{r t4 load model}
# load the model


plot(t4_rcs,type = "cumhaz", ci = FALSE, main="Cumulative hazards")
plot(t4_rcs,type = "hazard", ci = FALSE, main = "Hazards", max.time=10)
plot(t4_rcs,type = "survival", ci = FALSE, main = "Survival")

```

#  Transition 5: INT -> INT.CMD

```{r transition 5 rcs, eval  = TRUE}

ms_df<-reference_ms_df%>%mutate(baseline.a = int.a)


# Best fitting model number of knots

t5_rcs <- flexsurvspline(
  Surv(time = time, event = status) ~
   rcs(baseline.a) + gender + ethnicity  + IMD +smoking_status+icmm_pgs ,
  subset = (trans == 5),
  data = ms_df,
  k = t5_knots,
  scale = "hazard"
)

# Save the model
saveRDS(t5_rcs, file = "t5_rcs.rda")

```

```{r t5 load model}

plot(t5_rcs,type = "cumhaz", ci = FALSE, main="Cumulative hazards")
plot(t5_rcs,type = "hazard", ci = FALSE, main = "Hazards", max.time=10)
plot(t5_rcs,type = "survival", ci = FALSE, main = "Survival")

```

#  Transition 6: INT -> CVE


```{r transition 6 rcs, eval= TRUE}

ms_df<-reference_ms_df%>%mutate(baseline.a = int.a)


# Best fitting model number of knots

# Run the best fitting model

t6_rcs <- flexsurvspline(
  Surv(time = time, event = status) ~
   baseline.a + gender + ethnicity  + IMD +  smoking_status+icmm_pgs ,
  subset = (trans == 6),
  data = ms_df,
  k = t6_knots,
  scale = "hazard"
)

# Save the model
saveRDS(t6_rcs, file = "t6_rcs.rda")

```

```{r t6 load model}
# load the model


plot(t6_rcs,type = "cumhaz", ci = FALSE, main="Cumulative hazards")
plot(t6_rcs,type = "hazard", ci = FALSE, main = "Hazards", max.time=10)
plot(t6_rcs,type = "survival", ci = FALSE, main = "Survival")

```

# Transition 7: INT -> Death

```{r transition 7 rcs, eval = TRUE}
# Best fitting model number of knots
ms_df<-reference_ms_df%>%mutate(baseline.a = int.a)


t7_rcs <- flexsurvspline(
  Surv(time = time, event = status) ~
    baseline.a + 
    gender +
    ethnicity  +
    IMD + 
  
    smoking_status+icmm_pgs ,
  subset = (trans == 7),
  data = ms_df,
  k = t7_knots,
  scale = "hazard"
)


# Save the model
saveRDS(t7_rcs, file = "t7_rcs.rda")
```

```{r t7 load model}
# load the model


plot(t7_rcs,type = "cumhaz", ci = FALSE, main="Cumulative hazards")
plot(t7_rcs,type = "hazard", ci = FALSE, main = "Hazards", max.time=10)
plot(t7_rcs,type = "survival", ci = FALSE, main = "Survival")

```


# Transition 8: CMD-> CMD.INT


```{r transition 8 rcs, eval = TRUE}

# Best fitting model number of knots
ms_df<-reference_ms_df%>%mutate(baseline.a = cmd.a)


t8_rcs <- flexsurvspline(
  Surv(time = time, event = status) ~
  rcs(baseline.a) + gender + ethnicity  + IMD + smoking_status +icmm_pgs,
  subset = (trans == 8),
  data = ms_df,
  k = t8_knots,
  scale = "hazard"
)


# Save the model
saveRDS(t8_rcs, file = "t8_rcs.rda")

```

```{r t8 load model}
# load the model

plot(t8_rcs,type = "cumhaz", ci = FALSE, main="Cumulative hazards")
plot(t8_rcs,type = "hazard", ci = FALSE, main = "Hazards", max.time=10)
plot(t8_rcs,type = "survival", ci = FALSE, main = "Survival")


```


# Transition 9: CMD -> CVE


```{r transition 9 rcs, eval = TRUE}

ms_df<-reference_ms_df%>%mutate(baseline.a = cmd.a)

# Run the best fitting model

t9_rcs <- flexsurvspline(
  Surv(time = time, event = status) ~
    rcs(baseline.a) + gender + ethnicity  + IMD  + smoking_status+icmm_pgs ,
  subset = (trans == 9),
  data = ms_df,
  k = t9_knots,
  scale = "hazard"
)

# Save the model
saveRDS(t9_rcs, file = "t9_rcs.rda")

```

```{r t9 load model}
# load the model

plot(t9_rcs,type = "cumhaz", ci = FALSE, main="Cumulative hazards")
plot(t9_rcs,type = "hazard", ci = FALSE, main = "Hazards", max.time=10)
plot(t9_rcs,type = "survival", ci = FALSE, main = "Survival")

```

# Transition 10: CMD->Death

```{r rcs transition 10 , eval = TRUE}
# Best fitting model number of knots

ms_df<-reference_ms_df%>%mutate(baseline.a = cmd.a)


t10_rcs <- flexsurvspline(
  Surv(time = time, event = status) ~
   rcs(baseline.a) + gender + ethnicity  + IMD  + smoking_status +icmm_pgs,
  subset = (trans == 10),
  data = ms_df,
  k = t10_knots,
  scale = "hazard"
)


# Save the model
saveRDS(t10_rcs, file = "t10_rcs.rda")
```

```{r t10 load model}
# load the model


plot(t10_rcs,type = "cumhaz", ci = FALSE, main="Cumulative hazards")
plot(t10_rcs,type = "hazard", ci = FALSE, main = "Hazards", max.time=10)
plot(t10_rcs,type = "survival", ci = FALSE, main = "Survival")

```

# Transition 11: INT.CMD ->CVE

```{r rcs transition 11, eval = TRUE}
# Best fitting model number of knots

ms_df<-reference_ms_df%>%mutate(baseline.a = int.cmd.a)


t11_rcs <- flexsurvspline(
  Surv(time = time, event = status) ~
    baseline.a + gender + ethnicity  + IMD  + smoking_status +icmm_pgs,
  subset = (trans == 11),
  data = ms_df,
  k = t11_knots,
  scale = "hazard"
)

# Save the model
saveRDS(t11_rcs, file = "t11_rcs.rda")
```

```{r t11 load model}
# load the model


plot(t11_rcs,type = "cumhaz", ci = FALSE, main="Cumulative hazards")
plot(t11_rcs,type = "hazard", ci = FALSE, main = "Hazards", max.time=10)
plot(t11_rcs,type = "survival", ci = FALSE, main = "Survival")

```

# Transition 12: INT.CMD -> Death


```{r rcs transition 12, eval = TRUE}
# Run the best fitting model

ms_df<-reference_ms_df%>%mutate(baseline.a = int.cmd.a)


t12_rcs <- flexsurvspline(
  Surv(time = time, event = status) ~ 
     baseline.a + gender + ethnicity  + IMD  + smoking_status +icmm_pgs,
  subset = (trans == 12),
  data = ms_df,
  k = t12_knots,
  scale = "hazard"
)

# Save the model
saveRDS(t12_rcs, file = "t12_rcs.rda")

```

```{r t12 load model}
# load the model

plot(t12_rcs,type = "cumhaz", ci = FALSE, main="Cumulative hazards")
plot(t12_rcs,type = "hazard", ci = FALSE, main = "Hazards", max.time=10)
plot(t12_rcs,type = "survival", ci = FALSE, main = "Survival")

```


# Transition 13: CMD.INT -> CVE


```{r rcs transition 13, eval = TRUE}
# Best fitting model number of knots
ms_df<-reference_ms_df%>%mutate(baseline.a = cmd.int.a)


t13_rcs <- flexsurvspline(
  Surv(time = time, event = status) ~
    rcs(baseline.a) + gender + ethnicity  + IMD  + smoking_status +icmm_pgs,
  subset = (trans == 13),
  data = ms_df,
  k = t13_knots,
  scale = "hazard"
)

# Save the model
saveRDS(t13_rcs, file = "t13_rcs.rda")

```

```{r t13 load model}
# load the model



plot(t13_rcs,type = "cumhaz", ci = FALSE, main="Cumulative hazards")
plot(t13_rcs,type = "hazard", ci = FALSE, main = "Hazards", max.time=10)
plot(t13_rcs,type = "survival", ci = FALSE, main = "Survival")
```

# Transition 14: CMD.INT -> death

```{r transition 14 rcs, eval = TRUE}

# Run the best fitting model

ms_df<-reference_ms_df%>%mutate(baseline.a = cmd.int.a)


t14_rcs <- flexsurvspline(
  Surv(time = time, event = status) ~
    baseline.a+ 
    gender+
    ethnicity+
    IMD+
    smoking_status+
    icmm_pgs ,
  subset = (trans == 14),
  data = ms_df,
  k = t14_knots,
  scale = "hazard"
)

# Save the model
saveRDS(t14_rcs, file = "t14_rcs.rda")

```

```{r t14 load model}
# load the model

plot(t14_rcs,type = "cumhaz", ci = FALSE, main="Cumulative hazards")
plot(t14_rcs,type = "hazard", ci = FALSE, main = "Hazards", max.time=10)
plot(t14_rcs,type = "survival", ci = FALSE, main = "Survival")

```


